Variational oblique predictive clustering trees

### 2024-12-13 reports/paper/vspyct.tex
grammar fixes

This integral rarely yields a closed-form solution.
For cases with multiple dimensions, the computational burden often renders Monte Carlo methods infeasible.
The VB approach offers an approximation to this issue by closely mimicking the true posterior $p(\theta|x)$ with an approximate distribution $q_{\omega^*}(\theta)$. This variational distribution is part of a family of distributions $q_\omega(\theta)\in \mathcal{Q}$, parameterized by $\omega \in \Omega$, where $\Omega$ represents the potential latent parameter values.
Typically, the mean-field variational family is utilised, assuming independence among the parameters $\omega$~\citep{Blei_2017}. 
To find the optimal $\omega^*$, one minimises the \gls{kl} divergence $KL(q_\omega(\theta) \parallel p(\theta|x))$:

@@

    \Procedure{LearnSplit}{$\mathcal{D}, \theta, E, \lambda, \beta, \sigma$}
        \State Initialize $\Theta = \{\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w, \mu_b, \sigma_b^2\}$ \Comment{Initialize variational parameters}
        \State Initialize optimizer: $\alpha \gets \lambda$
        \For{$e \in \{1, \dots, E\}$} \Comment{Iterate over epochs}
            \For{each mini-batch $\mathcal{B} \subseteq \mathcal{D}$ of size $\beta$}
                \State Sample $\mathbf{w} \sim \mathcal{N}(\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w)$, $b \sim \mathcal{N}(\mu_b, \sigma_b^2)$
                \State Compute impurity $\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)$ as:
                \[
                \Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b) = \sum_{\text{right}} \rho(\mathbf{w}^\top \mathbf{x} + b) \cdot \text{Var}_\rho(\mathbf{Y}) + \sum_{\text{left}} (1 - \rho(\mathbf{w}^\top \mathbf{x} + b)) \cdot \text{Var}_\lambda(\mathbf{Y})
                \]
                \State Set the observed impurity value: $\Omega_\text{obs} = \frac{\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)}{2}$
                \State Condition on the observed impurity using:
                \[
                \log p(\Omega_\text{obs}|\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b))
                \]
                \State Compute ELBO $\mathcal{L}(\mathcal{B}; \Theta)$ as per~\eqref{eq:elbo}
                \State Compute gradient $\nabla_{\Theta} \text{ELBO}$
                \State Update $\Theta \gets \Theta + \alpha \nabla_{\Theta} \text{ELBO}$ \Comment{Gradient descent update}
            \EndFor
            \If{early stopping criterion satisfied}
                \State \textbf{break}
            \EndIf
        \EndFor
        \State \textbf{return} $\Theta$
    \EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\Function{MC\_Predict}{$\Omega$, $\mathbf{x}$, $M$}
    \State Initialize $\hat{y}_{\text{sum}} = 0$
    \For{$m = 1$ to $M$}
        \State $\hat{y}_{\text{sum}} \gets \hat{y}_{\text{sum}} + \text{Predict}(\Omega, \mathbf{x}, m)$
    \EndFor
    \State \Return $\hat{y} = \frac{\hat{y}_{\text{sum}}}{M}$
\EndFunction

@@

However, the overall feature importance is derived by aggregating these normalized ratios across all splits within the tree.
Even if \textit{salary} appears strongly influential at the root node, the final importance calculation considers both the magnitude and the stability of the weights across all nodes.
A feature consistently demonstrating a favorable \(\frac{\mathbb{E}[w]}{\text{Var}(w)}\) ratio throughout multiple splits will be deemed more important globally.
In this way, the final importance assessment accounts for both the strength and certainty of each feature’s contribution to the model’s decision-making process.

@@

For multi-target regression, a normalized Mean Absolute Error (\textit{NMAE}) was used to ensure comparability across targets with different value ranges. Letting \(y_{ij}\) and \(\hat{y}_{ij}\) be the true and predicted values for the \(j\)-th target of the \(i\)-th instance, and \(T\) be the total number of target variables, the \(\textit{NMAE}\) is defined as:
\[
\textit{NMAE} = \frac{1}{T} \sum_{j=1}^{T} \left( \frac{\frac{1}{N} \sum_{i=1}^{N} | y_{ij} - \hat{y}_{ij} |}{\max_i(y_{ij}) - \min_i(y_{ij})} \right)
\]
If \(\max_i(y_{ij}) - \min_i(y_{ij}) = 0\) for a given target \(j\), this denominator is replaced by 1, effectively leaving that target unscaled.

>>>>

This integral rarely yields a closed-form solution.
For cases with multiple dimensions, the computational burden often renders Monte Carlo methods infeasible.
The VB approach offers an approximation to this issue by closely mimicking the true posterior $p(\theta|x)$ with an approximate distribution $q_{\omega^*}(\theta)$. This variational distribution is part of a family of distributions $q_\omega(\theta)\in \mathcal{Q}$, parameterised by $\omega \in \Omega$, where $\Omega$ represents the potential latent parameter values.
Typically, the mean-field variational family is utilised, assuming independence among the parameters $\omega$~\citep{Blei_2017}. 
To find the optimal $\omega^*$, one minimises the \gls{kl} divergence $KL(q_\omega(\theta) \parallel p(\theta|x))$:

@@

    \Procedure{LearnSplit}{$\mathcal{D}, \theta, E, \lambda, \beta, \sigma$}
        \State Initialise $\Theta = \{\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w, \mu_b, \sigma_b^2\}$ \Comment{Initialise variational parameters}
        \State Initialise optimiser: $\alpha \gets \lambda$
        \For{$e \in \{1, \dots, E\}$} \Comment{Iterate over epochs}
            \For{each mini-batch $\mathcal{B} \subseteq \mathcal{D}$ of size $\beta$}
                \State Sample $\mathbf{w} \sim \mathcal{N}(\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w)$, $b \sim \mathcal{N}(\mu_b, \sigma_b^2)$
                \State Compute impurity $\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)$ as:
                \[
                \Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b) = \sum_{\text{right}} \rho(\mathbf{w}^\top \mathbf{x} + b) \cdot \text{Var}_\rho(\mathbf{Y}) + \sum_{\text{left}} (1 - \rho(\mathbf{w}^\top \mathbf{x} + b)) \cdot \text{Var}_\lambda(\mathbf{Y})
                \]
                \State Set the observed impurity value: $\Omega_\text{obs} = \frac{\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)}{2}$
                \State Condition on the observed impurity using:
                \[
                \log p(\Omega_\text{obs}|\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b))
                \]
                \State Compute ELBO $\mathcal{L}(\mathcal{B}; \Theta)$ as per~\eqref{eq:elbo}
                \State Compute gradient $\nabla_{\Theta} \text{ELBO}$
                \State Update $\Theta \gets \Theta + \alpha \nabla_{\Theta} \text{ELBO}$ \Comment{Gradient descent update}
            \EndFor
            \If{early stopping criterion satisfied}
                \State \textbf{break}
            \EndIf
        \EndFor
        \State \textbf{return} $\Theta$
    \EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\Function{MC\_Predict}{$\Omega$, $\mathbf{x}$, $M$}
    \State Initialise $\hat{y}_{\text{sum}} = 0$
    \For{$m = 1$ to $M$}
        \State $\hat{y}_{\text{sum}} \gets \hat{y}_{\text{sum}} + \text{Predict}(\Omega, \mathbf{x}, m)$
    \EndFor
    \State \Return $\hat{y} = \frac{\hat{y}_{\text{sum}}}{M}$
\EndFunction

@@

However, the overall feature importance is derived by aggregating these normalised ratios across all splits within the tree.
Even if \textit{salary} appears strongly influential at the root node, the final importance calculation considers both the magnitude and the stability of the weights across all nodes.
A feature consistently demonstrating a favourable \(\frac{\mathbb{E}[w]}{\text{Var}(w)}\) ratio throughout multiple splits will be deemed more important globally.
In this way, the final importance assessment accounts for both the strength and certainty of each feature’s contribution to the model’s decision-making process.

@@

For multi-target regression, a normalised Mean Absolute Error (\textit{NMAE}) was used to ensure comparability across targets with different value ranges. Letting \(y_{ij}\) and \(\hat{y}_{ij}\) be the true and predicted values for the \(j\)-th target of the \(i\)-th instance, and \(T\) be the total number of target variables, the \(\textit{NMAE}\) is defined as:
\[
\textit{NMAE} = \frac{1}{T} \sum_{j=1}^{T} \left( \frac{\frac{1}{N} \sum_{i=1}^{N} | y_{ij} - \hat{y}_{ij} |}{\max_i(y_{ij}) - \min_i(y_{ij})} \right)
\]
If \(\max_i(y_{ij}) - \min_i(y_{ij}) = 0\) for a given target \(j\), this denominator is replaced by 1, effectively leaving that target unscaled.
